A pandemic-scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-09-28 latex/main.tex
Rejig intro section on the ARG

Millions of SARS-CoV-2 genome sequences were collected in
response to the COVID-19 pandemic, forming a dataset of unprecedented
depth and richness.
Estimated genealogies are fundamental to
gleaning information from this ocean of data, 
and form the primary input to many downstream analyses.
A basic assumption of methods to infer genealogies from viral 
genetic sequence data is that recombination is negligible and 
the genealogy is a tree.
Recombinant lineages have risen to global prevalence in SARS-CoV-2,
and therefore such single-tree representations are necessarily
incomplete and potentially misleading.
We present sc2ts, a method to infer reticulate genealogies 
as an Ancestral Recombination Graph (ARG) for SARS-CoV-2
that automatically detects and incorporates recombination.  
We use sc2ts to infer an ARG of 2.48 million genomes,
and show that it captures a broad range of 
known evolutionary features and provides novel 
insights into the relationships between previously identified
recombinant lineages.
Together, the inference method, inferred ARG and 
suite of open-source analysis tools, 
represent a powerful resource for future research.

@@

The data collected during the COVID-19 pandemic as part of global
surveillance presents fundamental challenges to existing approaches
in two ways. Firstly, the sheer scale of data collected 
overwhelmed even approximate distance-matrix based methods.
At the peak in [month] 202X, there were over X thousand samples 
per day deposited in GISAID\cite{Shu2017}, with a total of 
over 17 million as of Sep 2025. % TODO check numbers
Regularly reinferring trees de-novo from such rapidly 
growing datasets is untenable.
The UShER\cite{Turakhia2021} method is based on adding 
samples to an existing tree using maximum parsimony, 
with periodic rearrangements to counter the inherent 
greediness of the approach\cite{DeMaio2023}.
The benefits of this simple model and the value of 
a comprehensive genealogy updated in real-time~\cite{McBroome2021}
were conclusively demonstrated, as the resource
quickly becoming an indispensible element 
of the pandemic response\cite{Hinrichs2024}.
The second challenge posed is that recombination is 
an essential element of SARS-CoV-2 
evolution\cite{Jackson2021,VanInsberghe2021,Focosi2022,
Gutierrez2022,Ignatieva2022,Tamura2023},
with multiple deep recombination events now known to be present in 
genealogy of the global population [cites??].
Although several methods have been proposed to \emph{detect}
recombinant samples~\cite{
Turakhia2022,%RIPPLES
Zhou2023virusrecom,% CovRecomb
Alfonsi2024,%RecombinHunt 
Li2024CovRecomb%CovRecomb 
},
this information is not incorporated into the phylogenies used 
in innumerable studies~\cite{McBroome2021,Hadfield2018},
and recombination continues to be treated as a 
nuisance factor.

@@

Sc2ts reconstructs an ARG by incrementally adding daily batches of 
sequences (Figure~\ref{fig:methodology}).
First, sc2ts infers likely ``copying paths'' connecting each sample in the 
batch to node(s) in the current ARG
(allowing recombination) using a simplified version of the Li and Stephens (LS) model
\cite{LiStephens2003}. 
% Important to go into a bit of detail about LS here as it's so fundamental
% and novel in the area.
The LS model is a Hidden Markov Model (HMM) widely used in human genomics
to approximate the effects of mutation and recombination on genetic
inheritance\citep{Delaneau2019-wl,Browning2018-nk},
typically parameterised by population-scaled mutation and
recombination rates varying along the genome~\cite{Donnelly2010-coalescent}.
In the interest of interpretability, we simplified this model to 
use a single parameter $k$ which essentially 
controls the number of recurrent mutations that is preferred to 
switching to a different parent via recombination (we use $k=4$ here).
We used the highly efficient LS HMM implementation in tsinfer\cite{Kelleher2019},
with some additional optimisations enabled by our simplified model,
to find exact Viterbi solutions of these copying paths for each sample
given the current ARG.
Note that this is a form of parsimony, with the recombination
penalty $k$ determining the relative parsimoniousness of recombinant
vs non-recombinant solutions to the HMM.
The per-sample copying paths produced by the HMM (most of which do not 
involve recombination) form clusters of samples within a batch, which 
are then resolved using standard phylogenetic techniques.
Finally, sc2ts attaches the phylogenies of the clusters to the current ARG, and
applies parsimony-improving heuristics to address issues introduced 
by the inherent greediness of this strategy.
Sc2ts is implemented in Python,
building on core PyData infrastructure~\cite{Harris2020array,Virtanen2020scipy,
Mckinney2010data,Lam2015numba} and bioinformatics 
packages\cite{Shirley2015efficient,Kunzmann2018biotite,Kunzmann2023biotite}.
See STAR Methods for details on all aspects of the method.

@@

Using sc2ts, we built an ARG from 2,482,157 samples from the Viridian v04
dataset\cite{Hunt2024}, collected between 2020-01-01 and 2023-02-20.
Inference took 21 days on a XX core server with 500 GB of RAM 
($\sim 5$ CPU years; Figure~\ref{fig:inference_resources}).
After postprocessing, the ARG 
contains 2,689,054 nodes with estimated dates, 
2,748,838 edges, and 2,285,344 mutations at 29,893 sites. 
Including sample IDs and the Pango and Scorpio designations computed by
Pangolin~\cite{OToole2021assignment} as metadata for each node,
the ARG is stored in a 32MiB tszip format file.
Of the nodes, 855 represent recombination events, 
resulting in 316 distinct trees along the genome (some
events occured at the same breakpoints).
% It's confusing to talk about recombinant QC here as we're talking 
% about the full ARG for the rest of the section.

The ARG is a comprehensive record of the evolutionary history of SARS-CoV-2 and 
is in close agreement with a state-of-the-art UShER phylogeny also built 
from the Viridian dataset \cite{Hunt2024}.
First, we validated the broad structure of the ARG against 
the UShER phylogeny \cite{Hunt2024}
by pruning both down to representative samples
for 1097 deeply-sampled lineages (STAR Methods).
Tanglegrams show that the ARG closely matches the phylogenetic
structure of the UShER tree (Figures~\ref{fig:tanglegram_base},
\ref{fig:tanglegram_delta},
\ref{fig:tanglegram_ba2},
\ref{fig:tanglegram_ba5}).
Next, we compared the sc2ts and UShER inferences 
from the Viridian dataset
by pruning both to their intersection of 2,475,418 samples and 27,507 sites,
% JK: this feels a bit wishy washy, but there really is no simple answer to
% this and I don't want to claim we're more parsimonious.
and found they were very similar in terms of parsimony
(Document~\ref{sec:supplement_sc2ts_usher_parsimony})
and imputation of missing and ambiguous bases
(Document~\ref{sec:supplement_sc2ts_usher_imputation}).
We also compared the mutational spectra of major VOCs in the ARG
against the mutational spectra from
the Viridian UShER phylogeny and
a previous study \cite{Bloom2023} (STAR Methods),
finding close agreement (Figure~\ref{fig:mutational_spectra}).

% FIXME this is too detailed, and too methodsy. Abstract out and summarise
% results
Input genome alignments are masked in sc2ts
so that deletions and ambiguous bases are treated as missing data by the HMM,
and once the samples are incorporated into the ARG,
these bases are imputed to carry the values inherited by the node in question.
Deletions are an important part of SARS-CoV-2 evolution \cite{Jeronimo2023}, however,
and to incorporate them into the ARG we identified a set of [XXX] sites
most frequently involved in deletions [cite source?]
and post-hoc mapped alignment data for these sites to the ARG using parsimony.
Although sites are treated independently in this approach,
the approximation successfully identified the most frequent deletions,
such as the recurrent 21991-21993 (See Document~\ref{sec:supplement_dels}
for additional analysis).
The node representing the origin of Alpha, with 290,407 descendants,
is correctly associated with key lineage-defining deletions:
11288-11296 (ORF1A del3675-3677),
21765-21770 (Spike del69-70),
21991-21993 (Spike del144), and
28271 (non-coding).
The Spike 69-70 deletion,
which causes S-Gene Target Failure \cite{Walker2021},
also emerges twice in the origin of Omicron -
in the node leading to the BA.1 variant (342,099 descendants),
and the node leading to the BA.4 variant (189,746 descendants).
To speed up inference and to avoid issues
caused by sites enriched for errors [citation?],
sc2ts masks out a set of 100 sites identified
as the most highly homoplasic
by running on a subset of the data (STAR Methods).
Of these sites, 9 are in the set of [1XX] ``problematic sites''
masked by UShER \cite{Hunt2024}.
In addition to remapping the data for sites identified
as being involved in deletions,
we added mutations for XX of these highly recurrent sites
that are used in Pango designation \cite{Rambaut2020} by post-hoc parsimony.
Each node in the ARG was then assigned
to one of 2058 Pango lineages using pangolin (STAR Methods).
All but 21,680 sampled sequences (0.87\%) agreed
with the source sequences in terms of Pango lineage assignment
(comparable to the 0.27\% disagreement between pangolin-data
versions 1.21 and 1.29 in the source Viridian metadata).
A majority of these lineages (1473 of 2058)
have monophyletic origins within the ARG,
increasing to 1779 (86\%) when we allow
for multiple sibling origination nodes
(Document~\ref{sec:supplement_pango_lineages}).

>>>>

Millions of SARS-CoV-2 genome sequences were collected in
response to the COVID-19 pandemic, forming a dataset of unprecedented
depth and richness.
Estimated genealogies are fundamental to
gleaning information from this ocean of data, 
and form the primary input to many downstream analyses.
A basic assumption of methods to infer genealogies from viral 
genetic sequence data is that recombination is negligible and 
the genealogy is a tree.
Recombinant lineages have risen to global prevalence in SARS-CoV-2,
and therefore such single-tree representations are necessarily
incomplete and potentially misleading.
We present sc2ts, a method to infer reticulate genealogies 
as an Ancestral Recombination Graph (ARG) for SARS-CoV-2
that automatically detects and incorporates recombination.  
We use sc2ts to infer an ARG of 2.48 million genomes,
and show that it captures a broad range of 
known evolutionary features and provides novel 
insights into the relationships between previously identified
recombinant lineages.
Together, the ARG, inference method and 
suite of open-source analysis tools, 
represent a powerful resource for future research.

@@

The data collected during the COVID-19 pandemic as part of global
surveillance presents fundamental challenges to existing approaches
in two ways. Firstly, the sheer scale of data collected 
overwhelmed even approximate distance-matrix based methods.
At the peak in [month] 202X, there were over X thousand samples 
per day deposited in GISAID\cite{Shu2017}, with a total of 
over 17 million as of Sep 2025. % TODO check numbers
Regularly reinferring trees de-novo from such rapidly 
growing datasets is untenable.
The UShER\cite{Turakhia2021} method is based on adding 
samples to an existing tree using maximum parsimony, 
with periodic rearrangements to counter the inherent 
greediness of the approach\cite{DeMaio2023}.
The benefits of this simple model and the value of 
a comprehensive genealogy updated in real-time~\cite{McBroome2021}
were conclusively demonstrated, as the resource
quickly became an indispensible element 
of the pandemic response\cite{Hinrichs2024}.
The second challenge posed is that recombination is 
an important force in SARS-CoV-2 
evolution\cite{Jackson2021,VanInsberghe2021,Focosi2022,
Gutierrez2022,Ignatieva2022,Tamura2023},
with multiple deep recombination events now known to be present in 
genealogy of the global population [cites??].
Although several methods have been proposed to \emph{detect}
recombinant samples~\cite{
Turakhia2022,%RIPPLES
Zhou2023virusrecom,% CovRecomb
Alfonsi2024,%RecombinHunt 
Li2024CovRecomb%CovRecomb 
},
this information is not incorporated into the phylogenies used 
in innumerable studies~\cite{McBroome2021,Hadfield2018},
and recombination continues to be treated as a 
nuisance factor.

@@

Sc2ts reconstructs an ARG by incrementally adding daily batches of 
sequences (Figure~\ref{fig:methodology}).
First, sc2ts infers likely ``copying paths'' connecting each sample in the 
batch to node(s) in the current ARG
(allowing recombination) using a simplified version of the Li and Stephens (LS) model
\cite{LiStephens2003}. 
% Important to go into a bit of detail about LS here as it's so fundamental
% and novel in the area.
The LS model is a Hidden Markov Model (HMM) widely used in human genomics
to approximate the effects of mutation and recombination on genetic
inheritance\citep{Delaneau2019-wl,Browning2018-nk},
typically parameterised by population-scaled mutation and
recombination rates varying along the genome~\cite{Donnelly2010-coalescent}.
In the interest of interpretability, we simplified this model to 
use a single parameter $k$ which essentially 
controls the number of recurrent mutations that is preferred to 
switching to a different parent via recombination (we use $k=4$ here).
We used the highly efficient LS HMM implementation in tsinfer\cite{Kelleher2019},
with some additional optimisations enabled by our simplified model,
to find exact Viterbi solutions of these copying paths for each sample
given the current ARG.
Note that this is a form of parsimony, with the recombination
penalty $k$ defining the relative parsimoniousness of recombinant
vs non-recombinant solutions to the HMM.
The per-sample copying paths produced by the HMM (most of which do not 
involve recombination) form clusters of samples within a batch, which 
are then resolved using standard phylogenetic techniques.
Finally, sc2ts attaches the phylogenies of the clusters to the current ARG, and
applies parsimony-improving heuristics to address issues introduced 
by the inherent greediness of this strategy.
Sc2ts is implemented in Python,
building on core PyData infrastructure~\cite{Harris2020array,Virtanen2020scipy,
Mckinney2010data,Lam2015numba} and bioinformatics 
packages\cite{Shirley2015efficient,Kunzmann2018biotite,Kunzmann2023biotite}.
See STAR Methods for details on all aspects of the method.

@@

Using sc2ts, we built an ARG from 2,482,157 samples from the Viridian v04
dataset\cite{Hunt2024}, collected between 2020-01-01 and 2023-02-20.
Inference took 21 days on a XX core server with 500 GB of RAM 
($\sim 5$ CPU years; Figure~\ref{fig:inference_resources}).
After postprocessing, the ARG 
contains 2,689,054 nodes,  % Note the estimated dates 
2,748,838 edges, and 2,285,344 mutations at 29,893 sites. 
Including sample IDs and the Pango and Scorpio designations computed by
Pangolin~\cite{OToole2021assignment} as metadata for each node,
the ARG is stored in a 32MiB tszip format file.
% It's confusing to talk about recombinant QC here as we're talking 
% about the full ARG for the rest of the section.
[Do we need to talk about recombination?]
Of the nodes, 855 represent recombination events, 
resulting in 316 distinct trees along the genome (some
events occured at the same breakpoints [vague, rephrase]).
[TODO note about loading up the time it takes to get the node 
data dataframes in a notebook?]

% PARA: we compared against the UShER tree in a bunch of ways
% NOTE: Trying to clarify that this of the pandemic period and we're 
% not trying to cover *all* history.
The ARG is a comprehensive record of the pandemic phase 
of SARS-CoV-2 evolutionary history, and accurately reflects
a range of results from the literature.
First, we compared the sc2ts ARG and with the state-of-the-art UShER 
tree inferred from the same dataset
by pruning both to their intersection of 2,475,418 samples and 27,507 sites
(STAR Methods)
% JK: this feels a bit wishy washy, but there really is no simple answer to
% this and I don't want to claim we're more parsimonious.
and found they were very similar in terms of parsimony
(Document~\ref{sec:supplement_sc2ts_usher_parsimony})
and imputation of missing and ambiguous bases
(Document~\ref{sec:supplement_sc2ts_usher_imputation}).
We also compared the mutational spectra of major VOCs in the ARG
against the UShER tree and 
a previous study \cite{Bloom2023} (STAR Methods),
finding close agreement (Figure~\ref{fig:mutational_spectra}).
We validated that the broad phylogenetic structure of the ARG and 
UShER tree closely match by generating tanglegrams for a set of representative
samples from 1097 deeply-sampled lineages (STAR Methods; 
Figures~\ref{fig:tanglegram_base},
\ref{fig:tanglegram_delta},
\ref{fig:tanglegram_ba2},
\ref{fig:tanglegram_ba5}).

% FIXME this is too detailed, and too methodsy. Abstract out and summarise
% results
Input genome alignments are masked in sc2ts
so that deletions and ambiguous bases are treated as missing data by the HMM,
and once the samples are incorporated into the ARG,
these bases are imputed to carry the values inherited by the node in question.

Deletions are an important part of SARS-CoV-2 evolution \cite{Jeronimo2023}, however,
and to incorporate them into the ARG we identified a set of [XXX] sites
most frequently involved in deletions [cite source?]
and post-hoc mapped alignment data for these sites to the ARG using parsimony.
Although sites are treated independently in this approach,
the approximation successfully identified the most frequent deletions,
such as the recurrent 21991-21993 (See Document~\ref{sec:supplement_dels}
for additional analysis).
The node representing the origin of Alpha, with 290,407 descendants,
is correctly associated with key lineage-defining deletions:
11288-11296 (ORF1A del3675-3677),
21765-21770 (Spike del69-70),
21991-21993 (Spike del144), and
28271 (non-coding).
The Spike 69-70 deletion,
which causes S-Gene Target Failure \cite{Walker2021},
also emerges twice in the origin of Omicron -
in the node leading to the BA.1 variant (342,099 descendants),
and the node leading to the BA.4 variant (189,746 descendants).
To speed up inference and to avoid issues
caused by sites enriched for errors [citation?],
sc2ts masks out a set of 100 sites identified
as the most highly homoplasic
by running on a subset of the data (STAR Methods).
Of these sites, 9 are in the set of [1XX] ``problematic sites''
masked by UShER \cite{Hunt2024}.
In addition to remapping the data for sites identified
as being involved in deletions,
we added mutations for XX of these highly recurrent sites
that are used in Pango designation \cite{Rambaut2020} by post-hoc parsimony.
Each node in the ARG was then assigned

to one of 2058 Pango lineages using pangolin (STAR Methods).
All but 21,680 sampled sequences (0.87\%) agreed
with the source sequences in terms of Pango lineage assignment
(comparable to the 0.27\% disagreement between pangolin-data
versions 1.21 and 1.29 in the source Viridian metadata).
A majority of these lineages (1473 of 2058)
have monophyletic origins within the ARG,
increasing to 1779 (86\%) when we allow
for multiple sibling origination nodes
(Document~\ref{sec:supplement_pango_lineages}).
